---
title: "Final Project"
author: "S. McGowan"
date: "`r Sys.Date()`"
output: 
  pdf_document: default
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#The above options names the code chunk as global-options, include=FALSE will hide this code chunk in final document. Knitr package will create report document
```

# Ideal Cardiovascular Health: An Analysis of Survey Data from Myanmar, 2021

## R Coding Final Project

## Simone McGowan

### Georgia Tech, Summer 2025

### GitHub Profile: "Simone-engineers" ; repository: "ECON-4001"

These data were obtained from the distributor, the **Inter-university Consortium for Political and Social Research**. It was collected in a survey format, to evaluate Metabolic Syndrome. The [information file](https://doi.org/10.3886/E146521V1-185521) in the hyperlink describes factor variable value assignments. The [raw data](https://doi.org/10.3886/E146521V1-185522) will be used to assess our outcome of interest, "ideal cardiovascular health" status based on metrics from the [American Heart Association](https://www.ahajournals.org/doi/10.1161/CIRCULATIONAHA.109.192703).

\newpage

# Table of Contents

1.  Literature Review & Motivation
2.  Data Description <!--subtopics have been indented below-->

-   Cleaning Data
    -   Midterm goals that were completed
    -   Midterm goals that were not completed +Next Steps
-   Analysis Goals
    -   Tables
    -   Visualizations

3.  Beyond Our R
4. Reference List

\newpage

# Literature Review

## Background

Metabolic syndrome is a reversible state comprised of having 3 of the 5 health conditions (elevated waist circumference, high triglycerides, high blood pressure, low HDL cholesterol, high fasting glucose) that puts a person at risk of cardiovascular disease (CD), stroke, and type 2 diabetes (Cleveland Clinic, 2023). Though the data was intended to evaluate risk of these three health outcomes, the purpose of the present analysis is to hone in on current cardiovascular health. In Myanmar, Stroke and Ischaemic Heart Disease were in the top 4 leading causes of death amongst men and women in 2021 (World Health Organization, n.d). It is therefore important to identify patterns in the population as a whole, as well as identify vulnerable populations, so that action can be taken on a community and policy/public health level (Lloyd-Jones et al., 2010). The presence of Ideal Heart health is linked with lower rates of mortality due to cardiovascular disease, as well as lower mortality rates due to diseases such as cancer. It is also helpful in reducing risk of cardiac events and stroke in pediatric populations later in life. An Ideal Heart health score will be created using ranges provided by the American Heart Association; relevant metrics include: BMI, total cholesterol, smoking status, physical activity, healthy diet, fasting plasma glucose, and blood pressure, with age determining cutoff values. (Lloyd-Jones et al., 2010)

## Socioeconomic status in Myanmar
According to the website Humanitarian Action the absolute poverty line is $2.15 per person per day in Myanmar, which is (2.15 USD per day x 4402.32 kyat/USD x 30 days) around 283949.64 kyats per person per month using today's exchange rate (OCHA, 2024). The socioeconomic status will be assessed by dividing monthly income of a person by number of children plus them self, to determine if they are at, above or below the poverty line. There is not enough information to assess marital contribution to household income.

\newpage

# Data Description
The data sources are linked in the abstract, coming from **Inter-university Consortium for Political and Social Research** hosted by the University of Michigan. The research was done by Su Su Maw from the University of Nursing in Yangon, Myanmar.

## Data Cleaning

```{r import,message=FALSE}
#load needed libraries and import raw data file
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)

setwd('C:/Users/smdot/OneDrive/Desktop/R-Summer25/Data/raw')
mydata=read.csv("MetS data to upload.csv", header=TRUE, stringsAsFactors =FALSE)
```

```{r removing variables}
#kept only variables in new data frame called new_data  
#rather than removing a lot of variables that were unusable upon examination or context
new_data_indeces=c(7,17,18,21,22,23,24,25,26,27,28,29,31,32,33,45,46,47,50,58)
new_data=mydata[new_data_indeces]
#confirmed that 20 variables that I want to keep remain.
#renaming variables and removing NULL values no longer applicable because 
#everything I have kept has a value and proper name
```

```{r creating social status variable}
#verify ChildrenNo and income are numeric
#monthly income/(number children+self) and rename column adjusted income.
class(new_data$ChildrenNo)
class(new_data$Income)
names(new_data)[1]="adjustedIncome"
new_data$adjustedIncome=new_data$adjustedIncome/(new_data$ChildrenNo+1)

```
```{r socioeconomic status, results='hide'}
# create a filter comparing adjusted income with the poverty line (283949.64 kyats/person). 
#If more than 10% greater Assign 3 for above, 1  for 10% below poverty line and 2 for those 
#around the poverty line (arbitrary plus or minus 10%). The new factor variable will be 
#called social status.

upperClassrows=which(new_data$adjustedIncome>1.1*283949.64)
belowPovertyrows=which(new_data$adjustedIncome<0.9*283949.64)
aroundPovertyrows=which(0.9*283949.64<=new_data$adjustedIncome &
new_data$adjustedIncome<=1.1*283949.64)
new_data$socialstatus=0
new_data$socialstatus[upperClassrows]=3
new_data$socialstatus[aroundPovertyrows]=2
new_data$socialstatus[belowPovertyrows]=1
as.factor(new_data$socialstatus)
```

```{r ideal heart health binary variables, results='hide'}
#create new dataframe with lifestyle binary variables only 
# it is a misnomer calling this new dataframe factors_df but keeping it
#switch values of VegFruit and Physical Exercise so that yes=1 and no=0 as with 
#other lifestyle binary variables

factors_df=new_data[,c(16,17,18,19,20)]
yesVegFruit=which(new_data$VegFruit==0)
noVegFruit=which(new_data$VegFruit==1)
factors_df$VegFruit[noVegFruit]=0
factors_df$VegFruit[yesVegFruit]=1

yesExercise=which(new_data$PhysicalExercise==0)
noExercise=which(new_data$PhysicalExercise==1)
factors_df$PhysicalExercise[noExercise]=0
factors_df$PhysicalExercise[yesExercise]=1

```

```{r ideal heart health scores temp variables,results='hide'}
#healthyChol, healthyBP, getsExercise, healthyEater, nonSmoker, 
#healthyFBS (fasting blood sugar), healthyBMI will have values of 1 if 
#they have met AHA health standards in that category. idealHeartscore will
#be the sum of these with a value of 0-7.

#healthy eating score calculation
healthyEater_filter=which(factors_df$VegFruit==1 & factors_df$Saltymeal==0 & factors_df$Oilymeal==0)
new_data$healthyEater=0
new_data$healthyEater[healthyEater_filter]=1

#healthy cholesterol calculation
healthyChol_filter=which(new_data$TotalCholesterol<200 & new_data$Age>20)
youngChol_filter=which(new_data$TotalCholesterol<170 & new_data$Age<=19)
new_data$healthyChol=0
new_data$healthyChol[healthyChol_filter]=1
new_data$healthyChol[youngChol_filter]=1

#healthy BP calculation
healthyBPfilter=which(new_data$SystolicBP<120 & new_data$DiastolicBP<80)
new_data$healthyBP=0
new_data$healthyBP[healthyBPfilter]=1

#healthy fasting blood sugar calculation
healthyFBS_filter=which(new_data$FBS<100)
new_data$healthyFBS=0
new_data$healthyFBS[healthyFBS_filter]=1

#BMI is metric used according to AHA citation (to address midterm comment)
#BMI calculation
healthyBMI_filter=which(new_data$BMI<25)
new_data$healthyBMI=0
new_data$healthyBMI[healthyBMI_filter]=1

#Never smoked status
nonSmoker_filter=which(factors_df$Pastsmoking==0)
new_data$nonSmoker=0
new_data$nonSmoker[nonSmoker_filter]=1

#Physical Exercise
getsExercise_filter=which(factors_df$PhysicalExercise==1)
new_data$getsExercise=0
new_data$getsExercise[getsExercise_filter]=1


#using dplyr to create new variable idealHealthscore
clean_data=new_data|>
  mutate(idealHealthscore=getsExercise+nonSmoker+
           healthyEater+healthyChol+healthyBMI+healthyBP+healthyBMI
)
```

## Midterm Goals Status Summary
I met some of the goals of this project, thank God I have to say. I put a lot on the midterm plan, and I learned later that there are easier ways to accomplish the same things. I had to make an assumption about age looking at the data. Though it was accounted for in total cholesterol filters, it was too complicated to ascertain "95th percentile" values for those youths age 18 and 19 for which BMI and blood pressure were impacted by age. They are not children, so I assumed they were close enough to group with the 20 and older adults to use the same cutoff values. The ideal heart score was calculated, as well as socioeconomic status factor variable created. I did not use a filter of sorts or loop to reassign 1's and 0's for binary variables. There were only 2 that I used that didn't agree with my system of yes=1 and no=0 when I manually looked through the information file. So I hardcoded those changes. I think it was excessive to plan to do several different types of visualizations in this time frame, so I am focusing on showing the regression for sleep versus ideal health rating, and a scatterplot for adjusted income versus ideal health score with trend lines for different socioeconomic status.


\newpage

## Summary Statistics for numeric variables of interest

```{r numeric variables}
summary_stats=summary(clean_data[c(1:15,29)])
summary_stats
```
\newpage 

## Analysis

You can also embed plots, for example:

```{r sleep regression
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
\newpage

#Beyond Our R
I did not use any packages beyond what we learned in class. I was inclined to use only base R, as I felt intimidated by using dplyr. However, I found that dplyr is a lot more efficient at creating new variables and manipulating their values with other columns in a data frame using the mutate function. What I learned is sufficient for my purposes, because I wanted find a relationship between socioeconomic status and ideal heart health, if there was one, to practice cleaning and analyzing health data as a biomedical engineering student. With the tools we learned this semester, I was able to get my hands dirty and gain skills I will put on my resume (cleaning, analysis, visualization of data). Regressions are used a lot in my field, and I didn't understand what they communicated until this course. Visualizing data used to take me hours to make a good plot, usually because of the details I tried to manually adjust in Excel (if I could find the setting). My design course professors really emphasized good graphics in our presentations of our projects, and I now feel I am proficient at basic R for my capstone project visualizations and data analysis in my last semester at Tech!

\newpage

#References
